Case study: Genome-Wide Association Analyses (GWAS)

### 2018-03-08 datasets/GWAS/GWAS.Rmd
Add covariate and effect size heatmaps

---
title: "GWAS datasets for FDR benchmarking"
author: "Keegan Korthauer"
date: "9/21/2017"
output: html_document
---

@@

### Summary

Here are the GWAS datasets we will use to benchmark the FDR methods. For each one
we provide the script to download the data and read into R. We'd like to verify 
that each dataset contains the desired inputs:

- p-value
- effect size
- standard error
- additional covariates (optional)

Note that while running all the bash code in this document will download the data, 
some of it doesn't seem to work properly with knitting (e.g. the ftp download 
in the second dataset), and I'm not committing 
the data itself to this repo. Run the bash code in the terminal separately to 
download the data yourself, and see the knitted `.html` document if you do not 
wish to download the datasets but still wish to see the output.

We'll also run the benchmark methods on each dataset and report the number of 
discoveries for each one. At first, we'll demo on just a __**small subset of 
SNPs**__, since the computation is time-intensive. After that is working 
smoothly, we'll scale it up to the full set of SNPs.

### Workspace setup

To run the benchmark methods on the GWAS datasets, we'll use some of the helper 
and wrapper functions from the simulated data analysis. We also create and source
an analgous version of the `sim_runner` function to apply to the real data.

```{r, wkspace-setup}
# Note that swfdr package depends on R version 3.4 or higher
setwd("/n/irizarryfs01_backed_up/kkorthauer/FDR")
source("benchmark-fdr/simulation_studies/R/simulation-helpers.R")
source("benchmark-fdr/datasets/R/evaluation-helpers.R")
datdir <- "/n/irizarryfs01_backed_up/kkorthauer/FDR/DATA/GWAS/"
resdir <- "/n/irizarryfs01_backed_up/kkorthauer/FDR/RESULTS/GWAS/"

# wrangling & plotting tools
library(data.table)
library(readxl)
library(dplyr)
library(ggplot2)

# benchmark methods
library(IHW)
library(ashr)
library(qvalue)
library(swfdr)
library(fdrtool)
```

### GWAS meta-analysis for BMI (included in Boca-Leek)

Here we download and analyze the GWAS BMI dataset analyzed in the Boca-Leek paper.

#### Data download

```{bash, GWAS1-download, eval=FALSE}
curl -O "http://portals.broadinstitute.org/collaboration/giant/images/3/3a/BMI.SNPadjSMK.zip"
```

Next we unzip the file, and delete the files we won't use (they provide results
subsetted by ancestry and sex), keeping only the file with all ancestries and
both sexes `BMI.SNPadjSMK.CombinedSexes.AllAncestry.txt`. Note that it seems that
the Boca-Leek paper actually used the 'EuropeanOnly' subset, but it's not clear
why (to me), and there's no reason we need to do the same.

```{bash, GWAS1-unzip, eval=FALSE}
unzip BMI.SNPadjSMK.zip
rm BMI.SNPadjSMK.zip
rm BMI.SNPadjSMK.Women.EuropeanOnly.txt
rm BMI.SNPadjSMK.Women.AllAncestry.txt
rm BMI.SNPadjSMK.Men.EuropeanOnly.txt
rm BMI.SNPadjSMK.Men.AllAncestry.txt
rm BMI.SNPadjSMK.CombinedSexes.EuropeanOnly.txt
```

@@

for 2,458,133 SNPs.

#### Running benchmark methods

First, we'll standardize the data frame to use the wrapper function. For now, we'll
assume that there is only a single 'independent covariate' for those methods that 
utilize this, and we'll use the sample size (number of subjects with that SNP
measured). **Still to-do: consider multiple covariates for methods that allow this.**

```{r, GWAS1-dat}
bmi <- bmi %>% 
        rename(pval = p_value,
               SE = stderr,
               effect_size = effect,
               ind_covariate = N) %>%
        mutate(zscore = effect_size/SE)

# Demo on a small subset of SNPs first (remove this later)
bmi <- bmi[sample(1:nrow(bmi), 100000),]
```

Now, we'll run the `run_benchmarks` helper function on the `bmi` data frame for
several alpha values and plot the results. This is fairly computationally intensive 
for some of the benchmark methods (since there are more than 2.5 million tests)
and best run on a machine with lots of memory.

```{r, GWAS1-run, results="hide"}
cutoffs <- c(0.01, 0.025, 0.05, 0.10, 0.20)
resfile <- paste0(resdir, "bmi_results_",
                  nrow(bmi), ".RData")
if (!file.exists(resfile)){
  t1 <- proc.time()
  bmi_results <- run_benchmarks(dat=bmi, alphas=cutoffs)
  save(bmi_results, file=resfile)
  message(paste0("Took ", round((proc.time()-t1)[3],1),
                 " seconds for ", nrow(bmi), " SNPs."))
}else{
  load(resfile)
}

# plot N rejects vs FDR cutoff
p <- ggplot(bmi_results, aes(alpha, n_rejects, color=method)) +
  geom_point() +
  geom_line() +
  xlab("FDR cutoff") + ylab("Number of rejections") +
  ggtitle("BMI GWAS Metanalysis with Sample Size Covariate") +
  theme_classic()
print(p)

# log-scale y axis
print(p + scale_y_continuous(trans="log2") + 
        ylab("Number of rejections (log2 scale)"))
```

And we'll clean up the workspace before moving on to the next dataset.

```{r, GWAS1-cleanup}
rm(bmi)
rm(bmi_results)
```

### GWAS for Asthma

[This study by Moffatt et al. (2010, NEJM)](http://www.nejm.org/doi/full/10.1056/NEJMoa0906312?query=TOC) performed a large
GWAS on  10,365 people with asthma and 16,110 controls 23 individual studies.

#### Data Download

Let's download the data from the Gabriel site. The second file contains metadata
for the columns in the main data file.

```{bash, GWAS2-download, eval=FALSE}
curl -O "https://beaune.cng.fr/gabriel/gabriel_results.zip"
curl -O "https://beaune.cng.fr/gabriel/gabriel_results_description.xls"
```

Next we unzip the file and remove the zipped copy.

```{bash, GWAS2-unzip, eval=FALSE}
unzip gabriel_results.zip
rm gabriel_results.zip
```

Next, we'll read in the unzipped `.txt` file and the results desription `.xls` 
file into R and verify that it contains
the necessary inputs to run the FDR benchmark comparisons.

```{r, GWAS2-verify-contents}
# read in meta-data
meta <- read_xls(paste0(datdir, "gabriel_results_description.xls"))
head(meta)
dim(meta)

# read in data
asthma <- fread(paste0(datdir, 
                "gabriel_asthma_meta-analysis_36studies_format_repository_NEJM.txt"),
                header=TRUE)
dim(asthma)

# verify meta-data rows = data cols
nrow(meta) == ncol(asthma)

# remove study-specific columns
empty <- which(!is.na(meta$description))
asthma <- asthma %>%
            select(empty)
meta <- meta %>% 
            slice(empty)

# remove SNPs with missing effect size/pvalues
asthma <- asthma %>% 
            filter(!is.na(P_fix) & !is.na(theta_fix))

dim(asthma)
head(asthma)
```

- `P_fix` and `P_ran`:p-values for fixed and random effect models
- `theta_fix` and `theta_ran`:effect size for fixed and random effect models
- `se_theta_fix` and `se_theta_ran`:standard error for fixed and random effect models
- additional covariates: 
    - `no_studies_meta`: Number of studies included in the meta-analysis with this SNP
    - `freq_all_1_max` and `freq_all_1_min`: Maximum and Minimum Allele Frequency of Allele 1 across samples

for 546,182 SNPs.

#### Running benchmark methods

First, we'll standardize the data frame to use the wrapper function. For now, we'll
assume that there is only a single 'independent covariate' for those methods that 
utilize this, and we'll use the sample size (number of studies with that SNP
measured). We'll also focus on the fixed effects tests first.
**Still to-do: consider multiple covariates for methods that allow this.**

```{r, GWAS2-dat}
asthma <- asthma %>% 
        rename(pval = P_fix,
               SE = se_theta_fix,
               effect_size = theta_fix,
               ind_covariate = no_studies_meta) %>%
        mutate(zscore = effect_size/SE)

# Demo on a small subset of SNPs first (remove this later)
asthma <- asthma[sample(1:nrow(asthma), 100000),]
```

Now, we'll run the `run_benchmarks` helper function on the `asthma` data frame for
several alpha values and plot the results. This is fairly computationally intensive 
for some of the benchmark methods
(since there are more than 500 thousand tests)
and best run on a machine with lots of memory.

```{r, GWAS2-run, results="hide"}
# use same cutoffs as for the bmi study (defined above)
resfile <- paste0(resdir, "asthma_results_",
                  nrow(asthma), ".RData")
if (!file.exists(resfile)){
  t1 <- proc.time()
  asthma_results <- run_benchmarks(dat=asthma, alphas=cutoffs)
  save(asthma_results, file=resfile)
  message(paste0("Took ", round((proc.time()-t1)[3],1),
                 " seconds for ", nrow(asthma), " SNPs."))
}else{
  load(resfile)
}

# plot N rejects vs FDR cutoff
p <- ggplot(asthma_results, aes(alpha, n_rejects, color=method)) +
  geom_point() +
  geom_line() +
  xlab("FDR cutoff") + ylab("Number of rejections") +
  ggtitle("Asthma GWAS Metanalysis with Sample Size Covariate") +
  theme_classic()
print(p)

# log-scale y axis
print(p + scale_y_continuous(trans="log2") + 
        ylab("Number of rejections (log2 scale)"))
```

And we'll clean up the workspace before moving on to the next dataset.

```{r, GWAS2-cleanup}
rm(asthma)
rm(asthma_results)
```

### GWAS for Lipid Levels

[This study by Willer et al. (2013, Nature Genetics)](http://www.nature.com/ng/journal/v45/n11/full/ng.2797.html) performed a large
GWAS on 188,577 and measured blood lipid levels (as well as several cardiovascular traits).

#### Data Download

Let's download the data from the University of Michigan site. The download page 
has a handy [metadata table](http://csg.sph.umich.edu//abecasis/public/lipids2013/)
that describes the contents of the columns in the results files.

There are several types of results files to download (in terms of which phenotype to analyze):

- LDL Cholesterol
- HDL Cholesterol
- Triglycerides
- Total Cholesterol

As well as two types of analyses:

- Joint analyses of GWAS and Metabochip
- analysis of Metabochip

We would like to look at the joint analysis of GWAS + Metabochip since we'd like 
to look at all available SNPs (not just those that have previously shown a possible
association with CVD). We could choose any or all of the traits, but I'll go ahead
and proceed with Triglycerides since I'm curious about the role of genetic variation 
on Triglyceride levels.

```{bash, GWAS3-download, eval=FALSE}
curl -O "http://csg.sph.umich.edu/abecasis/public/lipids2013/jointGwasMc_TG.txt.gz"
```

Next we uncompress the file.

```{bash, GWAS3-unzip, eval=FALSE}
gunzip jointGwasMc_TG.txt.gz 
```

Next, we'll read in the uncompressed `.txt` file into R and verify that it contains
the necessary inputs to run the FDR benchmark comparisons.

```{r, GWAS3-verify-contents}
# read in data
trig <- fread(paste0(datdir, "jointGwasMc_TG.txt"), header=TRUE)
dim(trig)
head(trig)
```

- `P_value`:p-value
- `beta`:effect size
- `SE`:standard error
- additional covariates: 
    - `N`: The number of individuals analyzed for this marker.
    - `Freq.A1.1000G.EUR`: Frequency of allele A1 from 1000G EUR sample.

for 2,439,432 SNPs.

#### Running benchmark methods

First, we'll standardize the data frame to use the wrapper function. For now, we'll
assume that there is only a single 'independent covariate' for those methods that 
utilize this, and we'll use the sample size (number of individuals with that SNP
measured). 
**Still to-do: consider multiple covariates for methods that allow this.**

Note that for this dataset, all reported effect sizes are positive. This is because
the alleles (A1 and A2) have been ordered such that the direction of effect are 
consistent. Since Scott assumes that the Z-scores are standard normal, we can't 
apply this method to a set of all positive Z-scores. To recover the direction of 
effect, we use the `Freq.A1.1000G.EUR` (allele frequency of A1) variable,
and use it to determine whether
`A1` (allele 1) is the reference allele or not. If the allele frequence of A1 is
greater than 0.5, we consider it the reference allele. Then we'll use the 
comparison of reference to minor as the positive direction. That is,
if A2 was reference,
then effect direction is negative. Otherwise it is positive. Since some of the
SNPs are missing the reference allele information, we cannot determine their 
direction of effect and they will be removed.

```{r, GWAS3-dat}
trig <- trig %>% 
        rename(pval = "P-value",
               SE = se,
               effect_size = beta,
               ind_covariate = N)%>%
        mutate(zscore = effect_size/SE)

# add effect size
trig <- trig %>% 
       mutate(A2.reference = Freq.A1.1000G.EUR < 0.5) %>%
       mutate(zscore = zscore * (1-2*A2.reference))

# remove SNPs with missing values for zscore (couldn't determine direction of 
# effect)
trig <- trig %>% 
  filter(!is.na(zscore))

# Demo on a small subset of SNPs first (remove this later)
trig <- trig[sample(1:nrow(trig), 100000),]
```

Now, we'll run the `run_benchmarks` helper function on the `trig` data frame for
several alpha values and plot the results. This is fairly computationally intensive 
for some of the benchmark methods
(since there are more than 500 thousand tests)
and best run on a machine with lots of memory.

```{r, GWAS3-run, results="hide"}
# use same cutoffs as for the bmi study (defined above)
resfile <- paste0(resdir, "trig_results_",
                  nrow(trig), ".RData")
if (!file.exists(resfile)){
  t1 <- proc.time()
  trig_results <- run_benchmarks(dat=trig, alphas=cutoffs)
  save(trig_results, file=resfile)
  message(paste0("Took ", round((proc.time()-t1)[3],1),
                 " seconds for ", nrow(trig), " SNPs."))
}else{
  load(resfile)
}

# plot N rejects vs FDR cutoff
p <- ggplot(trig_results, aes(alpha, n_rejects, color=method)) +
  geom_point() +
  geom_line() +
  xlab("FDR cutoff") + ylab("Number of rejections") +
  ggtitle("Triglyceride GWAS Metanalysis with Sample Size Covariate") +
  theme_classic()
print(p)

# log-scale y axis
print(p + scale_y_continuous(trans="log2") + 
        ylab("Number of rejections (log2 scale)"))

@@

And we'll clean up the workspace before moving on to the next dataset.

```{r, GWAS3-cleanup}
rm(trig)
rm(trig_results)
```

>>>>

---
title: "Case study: Genome-Wide Association Analyses (GWAS)"
author: "Keegan Korthauer"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
   html_document:
        toc: true
        toc_float: true
        highlight: tango
        number_sections: true
---

@@

# Summary

Here we will make use of the `SummarizedBenchmark` package to benchmark a 
Genome-Wide Association Study (GWAS) dataset. We will examine a GWAS 
meta-analysis for Body Mass Index (BMI), which was also included in the 
Boca-Leek manuscript.

# Workspace setup

This analysis requires that the software PLINK (version 1.9) 
is installed on your system.
PLINK is freely available for download 
[here](http://zzz.bwh.harvard.edu/plink/plink2.shtml).

```{r, wkspace-setup, results='hide', message=FALSE, warning=FALSE}
# data and results directories
system("mkdir -p DATA")
system("mkdir -p RESULTS")
datdir <- "DATA/"
resdir <- "RESULTS/"

# wrangling & plotting tools
library(data.table)
library(readxl)
library(readr)
library(dplyr)
library(ggplot2)
library(cowplot)

# comparison tools/functions
library(R.utils)
sourceDirectory("../R/")

# set up parallel backend
library(BiocParallel)
cores <- 8
multicoreParam <- MulticoreParam(workers = cores)
```

# Data preparation

We'll first download the GWAS summary statistics dataset from the 
GIANT consortium data portal.

## Data download

After downloading, we unzip the file, and delete the files we won't use (they provide results
subsetted by ancestry and sex), keeping only the file with European ancestry and
both sexes `BMI.SNPadjSMK.CombinedSexes.EuropeanOnly.txt`. Note that
the Boca-Leek paper also used the 'EuropeanOnly' subset, likely because using 
all ancestries would require that we assess the impact of population stratification
and adjust for it if present. For simplicity, we follow suit and use the homogeneous 
subset to avoid the impact of population stratification on our results.

```{r, GWAS1-download}
if(!file.exists(paste0(datdir,"BMI.SNPadjSMK.CombinedSexes.EuropeanOnly.txt"))){
  download.file(url="http://portals.broadinstitute.org/collaboration/giant/images/3/3a/BMI.SNPadjSMK.zip", 
             destfile=paste0(datdir,"BMI.SNPadjSMK.zip")) 
  unzip(paste0(datdir,"BMI.SNPadjSMK.zip"), exdir=datdir)
  file.remove(paste0(datdir,"BMI.SNPadjSMK.zip"))

  dfiles <- list.files(path=datdir, pattern="BMI.SNPadjSMK.*.txt", 
                     full.names = TRUE)
  dfiles <- dfiles[!grepl("BMI.SNPadjSMK.CombinedSexes.EuropeanOnly.txt", dfiles)]
  file.remove(dfiles)
}
```

We also download the 1000 genomes CEU reference population data, which will be used to 
determine which SNPs are LD buddies for the purposes of obtaining an independent set of SNPs.

```{r, GWAS1-refdata}
reffile <- paste0(datdir, "1000G_20101123_v3_GIANT_chr1_23_minimacnamesifnotRS_CEU_MAF0.01")
if (!file.exists(paste0(reffile, ".fam"))){
  download.file(url="http://neurogenetics.qimrberghofer.edu.au/iSECA/1000G_20101123_v3_GIANT_chr1_23_minimacnamesifnotRS_CEU_MAF0.01.zip", 
              destfile=paste0(reffile, ".zip"))
  unzip(paste0(reffile, ".zip"))
  file.remove(paste0(reffile, ".zip"))
}
```

@@

for `r nrow(bmi)` SNPs.

## Prune SNPs that are in Linkage Disequilibrium

Since nearby SNPs in linkage disequilibrium may not constitute separate discoveries, we carry out 
a pruning step. This commonly done by examining correlation of genotypes among nearby markers, but
since we don't have the genotype data, we'll use the linkage disequilibrium results of common SNPs and 
previously observed LD blocks. Specifically, we'll use the phase 3 data from 1000 genomes, restricted
to the CEU population (since this most closely matches our population of European ancestry). This 
reference data was downloaded in a previous code chunk. Here we read in the list of more than
9 million SNPs with data in 1000 genomes and create a list of SNPs that overlap our set. We'll write 
this list to a file that matches PLINK format.

```{r, GWAS1-snplist, eval=TRUE}
# load reference data 
onekg <- fread(paste0(datdir, "1000G_20101123_v3_GIANT_chr1_23_minimacnamesifnotRS_CEU_MAF0.01.bim"))$V2

# construct input list of SNPs that overlap - 97.5% overlap
bmi$BP <- as.numeric(unlist(lapply(strsplit(bmi$markername, ":"), function(x) x[[2]])))
write_delim(bmi %>% dplyr::rename(CHR=chromosome, SNP=rs_id, A1=allele_1, A2=allele_2) %>%
              mutate(P=0.5, F_A=Freq_Allele1_HapMapCEU, F_U=F_A, CHISQ=1, OR=1) %>% 
              select(CHR, SNP, BP, A1, F_A, F_U, A2, CHISQ, P, OR) %>%
              filter(SNP %in% onekg), 
            path=paste0(datdir, "rslist.assoc"), delim="\t")

ol <- sum(bmi$rs_id %in% onekg)/nrow(bmi)
rm(onekg)
```

It turns out that `r signif(ol*100, 3)` percent of SNPs in our set are included in the 1000 genomes 
reference data. Next we'll use the PLINK software to read in our list of SNPs and the 1000 genomes
linkage disequilibrium data to 'clump' the SNPs into independent sets of SNPs that have LD correlation (r squared)
less than 0.2 with all neighbors within 250 kilobases of distance. We write the results to a file called `plink_clump.clumped`. We have to jump out of R here since PLINK is a command line tool.

```{bash, clump, eval=FALSE}
# use PLINK clumping tool to obtain a set of independent SNPs by LD
plink \
--bfile DATA/1000G_20101123_v3_GIANT_chr1_23_minimacnamesifnotRS_CEU_MAF0.01 \
--clump DATA/rslist.assoc \
--clump-field P \
--clump-p1 0.9999 \
--clump-p2 0.9999 \
--clump-r2 0.2 \
--clump-kb 250 \
--out DATA/plink_clump
```

Next, we'll read in the clumped results and subset our data by our list of independent SNPs.

```{r, GWAS1-prune}
clump <- fread(paste0(datdir, "plink_clump.clumped"))

# subset to include only the SNPs in the ld.list
bmi <- bmi[bmi$rs_id %in% clump$SNP,]

rm(clump)
```

After pruning, we are left with for `r nrow(bmi)` approximately independent SNPs.

Before moving on, we'll rename the relevant columns so that they will use common terms
across the different datasets. We'll also add a 'test_statistic' column.

```{r, GWAS1-reformat}
bmi <- bmi %>% 
  dplyr::rename(pval = p_value,
                SE = stderr,
                effect_size = effect,
                ind_covar_N = N,
                ind_covar_AF = Freq_Allele1_HapMapCEU) %>%
  mutate(test_statistic = qnorm(exp(log(pval)-log(2)), lower.tail=FALSE) * sign(effect_size)) %>%
  select(pval, SE, effect_size, ind_covar_N, ind_covar_AF, test_statistic) 
```

# Data Analysis

## Differential Testing
Since we do not have access to the individual genotype information and since
our data already contains the results of performing hypothesis tests of
association of BMI with each SNP (including p-values and effect sizes), we 
don't need to carry out any additional testing. We move on to checking the
covariate diagnostics and adjusting for multiple comparisons.

## Covariate Diagnostics

Here we look to see if the covariates do indeed look informative. First we look at 
the sample size covariate.

### Covariate one: Sample Size

```{r, fig.width=4.5, fig.height=3.5, message=FALSE}
rank_scatter(bmi, pvalue="pval", covariate="ind_covar_N") + 
  ggtitle("Covariate 1: Sample Size")
```

```{r, fig.width=10, fig.height=3.2, message=FALSE}
strat_hist(bmi, pvalue="pval", covariate="ind_covar_N", maxy=2) 
```

### Covariate two: Minor Allele Frequency

Next we look at the allele frequency covariate.

```{r, fig.width=4.5, fig.height=3.5, message=FALSE}
rank_scatter(bmi, pvalue="pval", covariate="ind_covar_AF") +
  ggtitle("Covariate 1: Minor allele frequency")
```

```{r, fig.width=10, fig.height=3.2, message=FALSE}
strat_hist(bmi, pvalue="pval", covariate="ind_covar_AF", maxy=2)
```

## Multiple-Testing Correction

First, we'll create an object of `BenchDesign` class to hold the data and 
add the benchmark methods to the `BenchDesign` object.

```{r, GWAS1-benchdesign, message=FALSE}
bd <- initializeBenchDesign()
```

Now, we're ready to construct the `SummarizedBenchmark` object, which will run
the functions specified in each method (these are actually sourced in from the
                                        helper scripts).

### Covariate one: Sample size

First we'll include the sample size covariate.

```{r, GWAS1-sb, results="hide", message=FALSE}
resN <- paste0(resdir, "bmi_N_summarizedBenchmark_",
               nrow(bmi), ".RData")

duration <- NA
if (!file.exists(resN)){
  t1 <- proc.time()
  sbN <- bd %>% buildBench(data=(bmi %>% mutate(ind_covariate=ind_covar_N)), 
                          ftCols = c("ind_covar_N"),
                          parallel=TRUE, BPPARAM=multicoreParam)
  metadata(sbN)$data_download_link <- 
    "http://portals.broadinstitute.org/collaboration/giant/images/3/3a/BMI.SNPadjSMK.zip"
  saveRDS(sbN, file=resN)
  duration <- round((proc.time()-t1)[3]/60,1)
}else{
  sbN <- readRDS(resN)
}
```

```{r, echo=FALSE}
if(!is.na(duration)){
  message("This step took ", duration, " minutes for ", nrow(bmi), " SNPs using ",
          cores, " cores.")
}
```

### Covariate two: Minor Allele Frequency

Now, we'll repeat the multiple testing correction using the other covariate (AF):

```{r, GWAS1-sb2, results="hide", message=FALSE}
resAF <- paste0(resdir, "bmi_AF_summarizedBenchmark_",
nrow(bmi), ".RData")

duration <- NA
if (!file.exists(resAF)){
  t1 <- proc.time()
  sbAF <- bd %>% buildBench(data=(bmi %>% mutate(ind_covariate=ind_covar_AF) %>%
                                    filter(!is.na(ind_covariate))), 
                            ftCols = c("ind_covar_AF"),
                            parallel=TRUE, BPPARAM=multicoreParam)
  metadata(sbAF)$data_download_link <-    
    "http://portals.broadinstitute.org/collaboration/giant/images/3/3a/BMI.SNPadjSMK.zip"
  saveRDS(sbAF, file=resAF)
  duration <- round((proc.time()-t1)[3]/60,1)
}else{
  sbAF <- readRDS(resAF)
}
```

```{r, echo=FALSE}
if(!is.na(duration)){
  message("This step took ", duration, " minutes for ", nrow(bmi), " SNPs using ",
  cores, " cores.")
}
```

## Benchmark Metrics

Next, we'll add the default performance metric for q-value assays and 
plot the results. We'll start with covariate one.

### Covariate one: Sample Size

First, we have
to rename the assay to 'qvalue'.

```{r, GWAS1-metrics}
# rename assay to qvalue
assayNames(sbN) <- "qvalue"
sbN <- addDefaultMetrics(sbN)
```

Now, we'll plot the results.

```{r, GWAS1-plot, results="hide"}
# plot nrejects by method overall and stratified by covariate
rejections_scatter(sbN,
                   supplementary=FALSE) +
  ggtitle("Covariate 1: Sample size")

rejection_scatter_bins(sbN, covariate="ind_covar_N", bins=4,
                       supplementary=FALSE) +
  ggtitle("Covariate 1: Sample size")

# upset plot 
plotFDRMethodsOverlap(sbN, 
                      alpha=0.05, nsets=ncol(sb),
                      order.by="freq", decreasing=TRUE,
                      supplementary=FALSE)

# properties of method-exclusive discoveries
methods <- c("ashs", "lfdr", "ihw-a10", "bl-df03", "qvalue", "bh", "bonf",
             "scott-empirical", "scott-theoretical")
mcols(sbN)$ind_covariate <- mcols(sbN)$ind_covar_N
plotCovariateBoxplots(sbN, alpha=0.1, nsets=6, methods=methods, trans=log2, maxNum=3) +
  ggtitle("Covariate 1: Sample size")

@@

```{r, fig.width=8, fig.height=3.5}
mcols(sbN)$tstat <- bmi %>% filter(!is.na(ind_covar_N)) %>% pull(effect_size)
covariateHeatmap(sbN, alpha=0.05, covname="tstat")
covariateHeatmap(sbN, alpha=0.05, covname="ind_covar_N")
```

### Covariate two: Minor Allele Frequency

Next, we'll look at the performance metrics for the other covariate (minor
allele frequency).

```{r, GWAS1-metrics2}
# rename assay to qvalue
assayNames(sbAF) <- "qvalue"
sbAF <- addDefaultMetrics(sbAF)
```

Now, we'll plot the results.

```{r, GWAS1-plot2, results="hide"}
# plot nrejects by method overall and stratified by covariate
rejections_scatter(sbAF, supplementary=FALSE) +
  ggtitle("Covariate 2: Minor allele frequency")

rejection_scatter_bins(sbAF, covariate="ind_covar_AF", bins=4,
                       supplementary=FALSE) +
  ggtitle("Covariate 2: Minor allele frequency")

# upset plot 
plotFDRMethodsOverlap(sbAF, 
                      alpha=0.05, nsets=ncol(sbAF),
                      order.by="freq", decreasing=TRUE,
                      supplementary=FALSE)

# properties of method-exclusive discoveries
methods <- c("ashs", "lfdr", "ihw-a10", "bl-df03", "qvalue", "bh", "bonf",
             "scott-empirical", "scott-theoretical")
mcols(sbAF)$ind_covariate <- abs(mcols(sbAF)$ind_covar_AF-0.5)
plotCovariateBoxplots(sbAF, alpha=0.1, nsets=6, methods=methods, maxNum=3)+
  ggtitle("Covariate 2: Minor allele frequency")
```

```{r, fig.width=8, fig.height=3.5}
mcols(sbAF)$tstat <- bmi %>% filter(!is.na(ind_covar_AF)) %>% pull(effect_size)
covariateHeatmap(sbAF, alpha=0.05, covname="tstat")
covariateHeatmap(sbAF, alpha=0.05, covname="ind_covar_AF")
```

# Session Information

```{r}
sessionInfo()
```
